---
title: "The grammar of data visualization"
output: html_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

```{r, echo = FALSE}
source("R/common.R")
```

# Preamble

Nice resources:

* [ggplot2 online documentation](http://docs.ggplot2.org/current)
* [The graph section of R cookbook](http://www.cookbook-r.com/Graphs)

### Needed libraries for this tutorial:
```{r, message = FALSE, warning = FALSE, results='hide'}
library(tidyverse)  # This loads among other things the ggplot2-package
library(ggmap)
library(gridExtra)
library(maps)
library(mapdata)
```

Note if you have not installed these package, check the [Installing](installing.html)-documentation.

### Getting example data into R

For this tutorial we are going to use the following data (information on them found [here](http://www.hafro.is/~einarhj/data))

```{r}
minke <- read.csv("http://www.hafro.is/~einarhj/data/minke.csv",
                  stringsAsFactors = FALSE)
sau <- read.csv("http://www.hafro.is/~einarhj/data/sau-crfm-country-catches.csv",
                stringsAsFactors = FALSE)
iceland <- read.csv("http://www.hafro.is/~einarhj/data/iceland.csv",
                    stringsAsFactors = FALSE)
```

Just get a quick overview of the data we use the __glimpse__-function:

```{r}
glimpse(minke)
glimpse(sau)
glimpse(iceland)
```

# ggplot: Key components

ggplot has __three__ key components: 

1.  __data__,

2. A set of __aesthetic mappings__ [aes] for the variables in the data, and 

3. At least one layer which describes how to render each observation. Layers
   are usually created with a __geom__ function.

`r columns(1)`
```{r}
ggplot(data = minke, aes(x = age, y = length)) + 
  geom_point()
```

Here we have basically just created a point-plot where __age__ is plotted ("mapped") on the x-axis and __length__ on the y-axis. One can use different syntax resulting in the same outcome:

```{r, eval = FALSE}
ggplot(minke, aes(x = age, y = length)) + geom_point()
ggplot(minke, aes(age, length))         + geom_point()
ggplot()                                + geom_point(data = minke, aes(age, length))
ggplot(data = minke)                    + geom_point(aes(x = age, y = length))
ggplot(minke)                           + geom_point(aes(age, length))
minke %>% ggplot()                      + geom_point(aes(age, length))
```

One has quite a set of options were one can control colour, size, shape and alpha (the transparency of the points):
```{r}
ggplot(minke) + 
  geom_point(aes(age, length), colour = "red", size = 2, shape = 3, alpha = 0.4)
```

One can also stored the plot in an object for later use:
```{r}
p <- ggplot(minke) + geom_point(aes(age, length))
```
The above is useful if one wants to build a plot step-by-step or arrange many plots together for display (see below).

# Using the value of varible to control aesthetics

In the above cases the colour, size, shape and alpha levels are fixed values set outside the `aes`-function. We can however use the value of a variable in the dataset to control the visualization. Below is just a list of some things that can be set:

### aesthetic: colour
`r columns(2)`

One can add more aesthetics to the plot, e.g. if one wants to distinguish between __sex__ or the __area__ where the whale was caught:
```{r}
p <- ggplot(minke)
p + geom_point(aes(age, length, colour = sex))
p + geom_point(aes(age, length, colour = area))
```

Colours can be manually specified using `scale_colour_manual`:
```{r}
p <- ggplot(minke)
p + geom_point(aes(age, length, colour = sex)) +
  scale_colour_manual(values = c("orange","brown"))
p + geom_point(aes(age, length, colour = area)) +
  scale_colour_manual(values = c("green","red"))
```

In the above case __sex__ and __area__ had only two values, hence only two colours were specified.

### aesthetic: shape
`r columns(1)`
```{r}
p + geom_point(aes(age, length, shape = sex))
```

## Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Create a code that results in these plots:
`r columns(2)`
```{r, echo = FALSE}
p + geom_point(aes(age, length, colour = maturity))
p + geom_point(aes(age, length, shape = maturity))
```

</div>
</div>

### aesthetic: size

`r columns(1)`
```{r}
p + geom_point(aes(age, length, size = stomach.volume))
```

To reveal overlays:
`r columns(2)`
```{r}
p + geom_point(aes(age, length, size = stomach.volume), alpha = 0.6)
p + geom_point(aes(age, length, size = stomach.volume), alpha = 0.3, col = "red")
```

# facetting

A plot can be subsetted by using the `face_wrap`-function. Here we e.g. split the plot up into the two survey areas (North and South):
`r columns(1)`
```{r}
ggplot(minke) + 
  geom_point(aes(age, length, colour = sex)) + 
  facet_wrap(~ area)
```


<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create a code that results in this plot:
`r columns(1)`
```{r, echo = FALSE}
ggplot(minke) +
  geom_point(aes(age, length, colour = area)) + 
  facet_wrap(~ sex)
```

</div>
</div>

# Plotting plots side-by-side

If one wanted two different plots side-by-side one needs to store each plot as and object and then use the `grid.arrange`-function. E.g.:

```{r, fig.width = 10}
p1 <- p + geom_point(aes(age, length), colour = "blue")
p2 <- p + geom_point(aes(age, length, shape = sex))
grid.arrange(p1, p2, ncol = 2)
```

# layers

In the above cases we have only one type of a layer, the point layer. ggplot2 has of course myriads of layers. Below are some examples that give a brief overview of some other layers.

### layer: line

`r columns(2)`
```{r}
ggplot(sau, aes(year, total)) + geom_line()
```

You may have noticed that by default the plot area shown cover only the range of the data. In the above case one may have wanted to have the y-plot have a starting point at zero. Here we would need to use the `expand_limits`-function:
```{r}
ggplot(sau, aes(year, total)) + geom_line() + expand_limits(y = 0)
```

## layer: bar charts

We can create histograms for discrete data using the `geom_bar`-function:
`r columns(1)`
```{r}
ggplot(minke, aes(maturity)) + geom_bar()
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Modify the above code to generate the following:

```{r, echo = FALSE}
ggplot(minke, aes(maturity)) + geom_bar() + facet_wrap(~ area, nrow = 2)
```
</div>
</div>

## layer: histograms

We can create histograms for discrete data using the `geom_bar`-function. That function has also an argument for controlling the binwidth:

`r columns(2)`
```{r}
p <- ggplot(minke, aes(length))
p + geom_histogram()
p + geom_histogram(binwidth = 50)
```

One may want to get a histogram show the size distribution of e.g. the different sexes:
`r columns(2)`
```{r}
p + geom_histogram(aes(fill = sex))
```

It is quite hard to get a visualization of the length distribution of the Females here, so it may be better to split the plot into different panels:
`r columns(1)`
```{r}
p + geom_histogram(aes(fill = sex)) + facet_wrap(~ sex, ncol = 1)
```

## layer: frequency polygons

Instead of a histogram it may also be better to show things as frequency lines:

`r columns(2)`
```{r}
p + geom_freqpoly()
p + geom_freqpoly(aes(colour = sex), binwidth = 50)
```

## layer: points as jitters

Add a little random noise to the data to avoid over-plotting can be done using the `geom_jitter`-function:

`r columns(2)`
```{r}
p <- ggplot(minke, aes(sex, length))
p + geom_point()
p + geom_jitter()
```

## layer: summarise via boxplots or violins

Instead of a jitter plot, a more condensed way to show the distribution of the data is to use box- or violinplots:

`r columns(2)`
```{r}
p + geom_boxplot()
p + geom_violin()
```

## layer: overlay data and summary layers

In ggplot one can have more than one layer. E.g. in if we generate a summary distribution plot one may also want the get a "glimpse" at the raw data:

`r columns(1)`
```{r}
p + geom_boxplot() + geom_jitter(colour = "red", alpha = 0.3)
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create the following 3 layer plot:

```{r, echo = FALSE}
p + geom_violin() + geom_boxplot() + geom_jitter()
```

</div>
</div>

### layer: add smoother to a plot

Another example where we add layers could be adding a smoother to a point-plot:

`r columns(2)`
```{r}
p <- ggplot(minke, aes(age, length))
p + geom_point() + geom_smooth()
p + geom_point() + geom_smooth(span = 0.1)
```

We even have some specific models we could try, here a linear model:
`r columns(1)`
```{r}
p + geom_point() + geom_smooth(method = "lm")
```

# Some controls

If we want to put a plot into a report we may not like the default that is given but to refine the plot further. There are a number of functions in ggplot2 that allow us to take full control on the final outlook.

### labels

So far, the labels in the plot have been just the variable names in the dataframe. To specify the labels one can use the `labs`-function:
`r columns(2)`
```{r}
p <- ggplot(minke, aes(age, length, colour = sex)) + geom_point()
p
p + labs(x = "Age [year]", y = "Length [cm]", 
         colour = "Sex", title = "My minke plot",
         subtitle = "Based on survey data from 2003-2007")
```

## breaks

Controlling  which values appear as tick marks one can use:
`r columns(3)`
```{r}
p <- ggplot(minke, aes(age, length)) + geom_point() + labs(x = NULL, y = NULL)
p
p +
  scale_x_continuous(breaks = c(5, 10, 15, 20, 25, 30, 35, 40, 45))
p +
  scale_x_continuous(breaks = seq(5, 45, by = 5)) +
  scale_y_continuous(breaks = seq(500, 950, by = 50))
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create code that mimics the following plot:
`r columns(2)`
```{r, echo = FALSE}
ggplot(sau, aes(year, total)) + 
  geom_line() + 
  expand_limits(y = 0) +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10)) +
  scale_y_continuous(breaks = seq(0, 300, by = 25))
```
</div>
</div>

## limits

If you only want "zoom into" a data area one can use `xlim` or `ylim`:
`r columns(3)`
```{r}
p <- ggplot(minke, aes(maturity, length))
p + geom_jitter()
p + geom_jitter() + ylim(600, 800)
p + geom_jitter() + ylim(NA, 800) # setting only one limit
```

For discrete variables:
`r columns(1)`
```{r}
p + geom_jitter() + ylim(600,800) + xlim("immature","mature")
```

### warning

But be careful when using with summary statistics, e.g.:
`r columns(2)`
```{r}
p + geom_boxplot()
p + geom_boxplot() + ylim(600, 800)
```

This is because when you specify the `ylim` the data outside that range are filtered out completely from the plot-data. The remedy is to wrap the function inside the `coord_cartesian`-function:
`r columns(1)`
```{r}
p + geom_boxplot() + coord_cartesian(ylim = c(600, 800))
```

# A little gis

The object __island__ contains the Latitude and Longitude of the Icelandic shoreline. We could try to do a point or a line plot:

`r columns(2)`
```{r}
p <- ggplot(iceland, aes(lon, lat)) + labs(x = NULL, y = NULL)
p + geom_point()
p + geom_line()
```

The point plot shows something that is close to recognizable. However in the line plot data are rearranged such the line is drawn from the smallest x-value, to the next-smallest x-value and so on. The __island__ object actually has the data arranged in a specific order. If one want to retain that order one needs to use the `geom_path`-function. In addition, because the data are in coordinates we want to set the correct aspect ratio between the x- and the y-axis, hence here we also need to use the `coord_map`-function
```{r}
p + geom_path()
p + geom_path() + coord_map()
```

So, gis-mapping is nothing more than having data in a structured order and one connect the dots with a line in addition to specifying the map projection.

## On maps

* Maps as background for r-plot can come from myriad of sources
* In this course we will largely focus on objects available in the map- and mapdata-packages.
* To get the data into ggplot2 friendly form (data.frame) we use the `map_data` function.

`r columns(1)`
```{r}
m <- map_data("world")
str(m)
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap()
```

The resolution of "world" is not very high as can be seen if only want Barbados:
```{r}
m <- map_data("world", region = "Barbados")
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap()
```

If we want higher resolution the object "worldHires" is often sufficient:
```{r}
m <- map_data("worldHires")
m <- m[m$long > -85 & m$long < -60 & m$lat > 0 & m$lat < 30,]
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap(ylim = c(7, 28))
```

But it is still not really good, e.g. if we check out St. Vincent:
```{r}
m <- m[m$region == "Saint Vincent",]
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap()
```

we only get the main island, not the Grenadines.

## gis: Lets generate a base map:

`r columns(1)`
```{r}
m <- ggplot(iceland, aes(lon, lat)) +
  theme_bw() +
  geom_polygon(fill = "grey90") +
  coord_map() +
  labs(x = NULL, y = NULL)
m
```

## gis: Add layers
`r columns(2)`
```{r}
m + geom_point(data = minke, aes(lon, lat))
m + geom_point(data = minke, aes(lon, lat, colour = area))
```

```{r}
m + geom_point(data = minke, aes(lon, lat, colour = sex))
m + geom_point(data = minke, aes(lon, lat, colour = year))
m + geom_point(data = minke, aes(lon, lat, colour = factor(year)))
m + geom_point(data = minke, aes(lon, lat, size = length), alpha = 0.2)
# possible remedy of the above plot - is beyound the basic introduction
m + geom_point(data = minke, aes(lon, lat, colour = length)) +
  scale_colour_gradient(low = "yellow", high = "red")
```

## gis: Another type of base map:

```{r}
m2 <- get_map(location = c(-19,65), zoom= 6, maptype = "satellite", source = "google")
m2 <- ggmap(m2) +
  labs(x = NULL, y = NULL)
m2
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Repeat previous plots where we use "iceland" or do new ones, but using the Google base as a background. E.g.:

`r columns(1, 1, 1.2)`
```{r, echo = FALSE}
m2 + geom_point(data = minke, aes(lon, lat, colour = maturity))
```


</div>
</div>

