---
title: "Intalling packages"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

## Installing packages

Packages (libraries) can be thought of an exention of the basic R. These are typically a set of functions or some data. Although the basic R comes with a number of standard packages they only form a tiny part of the current R-environment.

Generally packages are install with the following command:
```{r}
install.packages("packagename")
```

E.g. if you want to install the package __readr__ one would do:
```{r}
install.packages("readr")
```

When you use `install.packages` function you will see a bunch of text that appears in the console window. If you have installed a package successfully a text near the bottom should state something like:
```
* DONE (readr)

The downloaded source packages are in
	‘/tmp/RtmpVpJbb8/downloaded_packages’
```

If something else appears, then the package most likely did not install sucessfully. The problems may be:

* Your R-program is too old for the package
* Some additional programs may be needed 
* ...

Trying to decipher what may be the problem is a challenge for the novice. 

For the work ahead it is recomended that the following packages are installed:

* __ggplot2__: For data visualization.
* __dplyr__: For data manipulation.
* __tidyr__: For data tidying.
* __readr__: For data import.
* __readxl__: For reading data from Excel
* __lubridate__: Working with date and time
* __stringr__: Working with strings

This can be done by typing in the following R-code (run only one line at a time and watch closely the message that appear in the console).

```{r}
install.packages("ggplot2")
install.packages("dplyr")
install.packages("tidyr")
install.packages("readr")
install.packages("readxl")
install.packages("lubridate")
install.packages("stringr")
```


