---
title: "Intalling packages"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

## Installing packages

Packages (libraries) can be thought of an extensions of the basic R. These are typically a set of functions or some data. Although the basic R comes with a number of standard packages they only form a tiny part of the current R-environment.

Generally packages are installed with the following command:
```{r}
install.packages("packagename")
```

E.g. if you want to install the package __readr__ one would do:
```{r}
install.packages("readr")
```

When you use `install.packages` function you will see a bunch of text that appears in the console window. If you have installed a package successfully a text near the bottom should state something like:
```
* DONE (readr)

The downloaded source packages are in
	‘/tmp/RtmpVpJbb8/downloaded_packages’
```

If something else appears, then the package most likely did not install successfully. The problems may be:

* Your R-program is too old for the package
* Some additional programs may be needed 
* ...

Trying to decipher the error output that gets printed in the console may be a challenge for the novice.

One normally installs packages only once or when one wants to get the latest update. To see what packages are installed one can use the following function:
```{r, eval = FALSE}
installed.packages()
```

In RStudio one can obtain the same information by opening the "Packages" tab in the lower right window panel.

For the work ahead it is recommended that the following __tidyverse__ packages are installed:

* __ggplot2__: For data visualization
* __tibble__: Data frame wrapper
* __tidyr__: For data tidying
* __readr__: For data import
* __dplyr__: For data manipulation

This can be done by typing in the following R-code:
```{r}
install.packages("tidyverse")
```

If you don't get something like:
```
* DONE (tidyverse)
```
in one of the bottom lines in the console, then not all packages may have installed. So try as an alternative:
```{r}
install.packages("ggplot2")
install.packages("tibble")
install.packages("tidyr")
install.packages("readr")
install.packages("dplyr")
```

Besides the above packages you are advised to install also the following packages:

* __readxl__: For reading data from Excel
* __lubridate__: Working with date and time
* __stringr__: Working with strings

Installation of other packages may be called for later in this space.
