---
title: "More grammar"
output: 
  html_document: 
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

# Preamble

> NOTE: This document is only partially completed

In this exercise the power of combining functions from the dplyr and ggplot2 packages (and then some) as a data exploration tool to reveal potential patterns and trends is demonstrated.

It is expected that readers are already familiar with the basics of data [manipulation](dplyr.html) and [plotting](ggplot2.html). We will use the flying fish data again.

__Needed packages__:
```{r}
library(tidyverse)
library(lubridate)
```

__Getting the example data into R__:

```{r}
ff <- read.csv("http://www.hafro.is/~einarhj/crfmr/data-raw/flyingfish.csv",
               stringsAsFactors = FALSE)
```

Lets see what we got:
```{r}
glimpse(ff)
```

So we have `r nrow(ff)` observations. Take note that the variable type of Year, Month, Weight..kg. and Trips are numerical values (labelled as `<dbl>` above) as expected.
The column names are not to my liking (I want to minimize keyboard work in the code that follows) so I change the column names:
```{r}
names(ff) <- c("year", "month", "country", "vessel", "catch", "effort")
```

And because we are going to be most interested in the catch per unit effort data we might as well generate that variable:

```{r}
ff <-
  ff %>% 
  mutate(cpue = catch/effort)
```

# A tiny introduction to "date" format in R

There are two columns that refer to time, year and month. Because I want later to plot data along time I might as well set up the proper date variable. Here I use the `ymd` function from the lubridate package (for further reading check the introductory [vignette](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)). Since there is no specific "day" in the data I just use the first day of each month as a dummy constant.

```{r}
ff <- 
  ff %>% 
  mutate(cpue = catch/effort,
         date = ymd(paste(year, month, 1, sep = "-")))
```

If you now take a glimpse at the data you notice that the type for the date column is labelled as `<date>` (i.e. as expected).

# An overall view of the data

```{r}
ff %>% 
  ggplot(aes(date, effort, colour = country)) +
  geom_line() +
  facet_wrap(~ vessel) +
  labs(x = NULL, y = "Number of trips")
ff %>% 
  ggplot(aes(date, catch, colour = country)) +
  geom_line() +
  facet_wrap(~ vessel) +
  labs(x = NULL, y = "Catch [kg]")
ff %>% 
  ggplot(aes(date, cpue, colour = country)) +
  geom_line() +
  facet_wrap(~ vessel) +
  labs(x = NULL, y = "CPUE [kg/trip]")
```

The main patterns one observes is:

* There is strong seasonality in the data.
* The catch per trip within dayboats seems to be higher for Tobago than the other two countries.
* The catch per trip is generally higher for the iceboats than the dayboats.

To reveal long term trend in catch per unit of effort we could try to use a smoother:

```{r}
ff %>% 
  ggplot(aes(date, cpue, colour = country)) +
  geom_point(size = 0.5) +
  geom_smooth(span = 0.3) +
  facet_wrap(~ vessel) +
  labs(x = NULL, y = "CPUE [kg/trip]")
```

Among the general patterns are:

* The cpue of the Iceboats shows a general decline with time until around 2015, thereafter increasing to historical high values in the terminal year.
* The cpue of the Tobago dayboat show a general increase until around 2003 but thereafter some decline.
* There seems to be an increase in cpue of the dayboats in Barbados in recent years.
* The pattern in the St. Lucia data are characterized by a temporal increase in the beginning of this century.

In order to make comparisons across the cpue series one can normalize the data (notice that I start filtering the data such that only common year ranges in the data (1995-2007) are included in the analysis) such that mean of the data within groups (country and vessel) is the equal to 1 by using the combination of `group_by` and `mutate` functions:

```{r}
ff %>% 
  filter(year %in% 1995:2007) %>% 
  group_by(country, vessel) %>% 
  mutate(mean = cpue/mean(cpue)) %>% 
  ungroup() %>% 
  mutate(group = paste(country, vessel)) %>% 
  ggplot(aes(date, mean, colour = group, fill = group)) +
  theme_bw() +
  geom_hline(yintercept = 1) +
  geom_smooth(aes(y = mean), span = 0.3) +
  #facet_wrap(~ vessel + country) +
  labs(x = NULL, y = "CPUE index") +
  scale_fill_brewer(palette = "Set1") +
  scale_colour_brewer(palette = "Set1")
```

Here we have not plotted the actual data, just the "loess" smoother. We added a horizontal line (using `geom_hline`) to indicate the mean within each time series. The default ggplot colour scheme was also overwritten using the `scale_xxx_brewer` (type `?scale_fill_brewer` to get further information of the function) to set of functions.

# The seasonal patterns

> TODO: Add text to explain the code

```{r}
p <-
  ff %>% 
  filter(year %in% 1995:2007) %>%
  ggplot(aes(factor(month), cpue)) +
  geom_boxplot()
p
p + facet_wrap(~ vessel)

ff %>% 
  filter(year %in% 1955:2007) %>% 
  group_by(month, vessel) %>% 
  summarize(mean = mean(cpue),
            p005 = quantile(cpue, 0.05),
            p050 = quantile(cpue, 0.50),
            p095 = quantile(cpue, 0.95)) %>% 
  ggplot(aes(factor(month))) +
  geom_linerange(aes(ymin = p005,
                     ymax = p095)) +
  geom_point(aes(y = p050)) +
  geom_point(aes(y = mean), colour = "red") +
  facet_wrap(~ vessel, scale = "free_y")

ff %>% 
  filter(year %in% 1955:2007) %>% 
  ggplot(aes(factor(month), cpue)) +
  theme_bw() +
  stat_summary(fun.data = "mean_cl_boot", colour = "red", size = 1) +
  facet_wrap(~ vessel, scale = "free_y")
```

# The patterns among fleet

> TODO: Add text to explain the code

```{r}
ff %>% 
  filter(year %in% 1955:2007) %>% 
  ggplot(aes(vessel, cpue)) +
  theme_bw() +
  stat_summary(fun.data = "mean_cl_boot", colour = "red", size = 1)
```
