---
title: "Tidy data"
output: html_document
---

## Preamble

Data are often not in tidy format ....

Note that when we here refer to tidy data it is not the same thing as "data cleaning". The latter deals specifically with correcting typos.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

```{r}
library(tidyverse)
```

## Gather

Datasets are often arranged in a wide format, mainly because it is more readable for humans. Take e.g. this dataset representing catches by gear by year:
```{r, echo = FALSE}
catch <- data_frame(gear = rep(c("hook", "line", "nets"), each = 2),
                 year = rep(c(2001, 2002), 3),
                 catch = c(500, 750, 400, 600, 20, 20)) %>% 
  spread(year, catch)
```

```{r}
catch
```

Here the column names represent a variable name (year) and it contains the actual value (2001 and 2002). And in each row we have observation of catches in two rows not one.

To tidy dataset such as these we use the `gather`-function:

```{r}
catch %>% 
  gather(key = year, value = catch, `2001`:`2002`)
```

Here we selected the year columns and give it the (key) variable name "year" and then the actual values are given the variable name "catch".

Note that we could also have used a negative value for the selection of column:
```{r, eval = FALSE}
# not run
catch %>% 
  gather(key = year, value = catch, -gear)
```

In the above we basically select all columns (to gather) except the species column.

## Spread

`spread` is the exact opposite of `gather`, i.e. takes the data from a long format to a wide format.

```{r, echo = FALSE}
catch <- data_frame(gear = rep(c("hook", "line", "nets"), each = 2),
                 year = rep(c(2001, 2002), 3),
                 value = c(500, 750, 400, 600, 20, 20)) %>% 
  mutate(type = "catch")
effort <- 
  data_frame(gear = rep(c("hook", "line", "nets"), each = 2),
                 year = rep(c(2001, 2002), 3),
                 value = c(100, 150, 200, 300, 50, 25)) %>% 
  mutate(type = "effort")
df <- 
  bind_rows(catch, effort) %>% 
  select(gear, year, type, value) %>% 
  arrange(gear, year, type)
```

```{r}
df
```

In the example above an observation is a gear in a year, but each observation is spread across two rows, where the numerical value have different meanings. We can spread these values across the two types as follows:
```{r}
df %>% 
  spread(type, value)
```

## Separate

... to be added

